WIDS 2024 Final Project

### 2024-01-10 Code/R FIle.qmd
cleaned up package loading, data loading, and data cleaning

# useful git commands

**This command will update all the branches on your computer** git remote update origin --prune

showing off branch

test

### Loading Packages

```{r}
library(tidyverse)
library(usmap)
```

### Gathering Data

```{r}
# Union Data
union_wages <- read_rds("../Data/union_wages.rds")
union_states <- read_rds("../Data/union_states.rds")

# Election Data
house <- read_rds("../Data/house.rds")

# Wage Data
income_distribution <- read_rds("../Data/income_distribution.rds")

# Minimum Wage Data
minimum_wage <- read_csv("../Data/minimum_wage_data.csv")
```

### Cleaning Data

#### Examining data shape

First, lets get an idea of what our data looks like.

```{r}
# look at columns
head(union_wages)
head(union_states)
head(house)
head(income_distribution)
head(minimum_wage)
```

@@

#### Cleaning data script

@@

```{r}
income_distribution <- income_distribution |> 
  select(-number) |> # removing unwanted columns
  select(!ends_with("moe")) |> 
  select(year:income_mean) |> 
  distinct(year, race, income_median, income_mean, .keep_all = TRUE) |> # only include one line per year. needed bc of multiple sectors that are not being used in data analysis
  rename(median_income = income_median,
         mean_income = income_mean) # renaming columns
```

@@

### Combining dataframes

@@

#### State Dataframes

@@

```{r}
summary(full_state_data)
```

#### National Dataframes

##### Creating Adjusted Wages Columns

@@

```{r}
full_ntl_data <- income_distribution |> 
  left_join(union_wages, by = "year") |>
  left_join(state_to_ntl_adjusted, by = "year") |>
  left_join(state_to_ntl_employment, by = "year") |>
  mutate(decade = paste0((year - year %% 10), "s"), .after = year) |>
  mutate(mean_median_income_difference = mean_income - median_income, .after = mean_income) |>
   mutate(median_income_2020_adj = median_income * prop_2020_adj,
         mean_income_2020_adj = mean_income * prop_2020_adj,
         mean_median_income_difference_2020_adj = mean_median_income_difference * prop_2020_adj,
         wage_2020_adj = wage * prop_2020_adj,
         union_wage_2020_adj = union_wage * prop_2020_adj,
         nonunion_wage_2020_adj = nonunion_wage * prop_2020_adj)
```

@@

### EDA

```{r}
summary(full_ntl_data)
summary(full_state_data)
```

@@

```{r}
library(gganimate)
library(gifski)
library(transformr)
```

>>>>

# 1. Comparing interactions between collective bargaining rates, wages, and House of Representative elections

** Insert abstract of project, where we got our data, what our data looks like, and other information here

# 2. Data Analysis 
## 2.1 Getting Set Up

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(usmap) # Used for map plots
```

```{r}
# Union Data: obtained from Tidy Tuesday
union_wages <- read_rds("../Data/union_wages.rds")
union_states <- read_rds("../Data/union_states.rds")

# Election Data: Obtained from Tidy Tuesday
house <- read_rds("../Data/house.rds")

# Wage Data: Obtained from Tidy Tuesday
income_distribution <- read_rds("../Data/income_distribution.rds")

# Minimum Wage Data: Obtained from U.S. Department of Labor
minimum_wage <- read_csv("../Data/minimum_wage_data.csv",show_col_types = FALSE)
```

## 2.2 Basic Data Cleaning
### 2.2.1 Examining data shape

Before we do anything to our data, we will examine our data to ensure everything loaded in correctly

```{r}
head(union_wages)
head(union_states)
head(house)
head(income_distribution)
head(minimum_wage)
```

@@

### 2.2.2 Cleaning data script

@@

```{r}
income_distribution <- income_distribution |> 
  select(-number) |> # removing unwanted columns
  select(!ends_with("moe")) |> 
  select(year:income_mean) |> 
  distinct(year, race, income_median, income_mean, .keep_all = TRUE) |> # only include one line per year. needed bc of multiple sectors that are not being used in data analysis
  rename(median_income = income_median,
         mean_income = income_mean) |> # renaming columns
  filter(race == "All Races") |> 
  select(-race)
```

@@

### 2.2.3 Merging dataframes

@@

#### 2.2.3.1 State Dataframes

@@

```{r}
head(full_state_data)
summary(full_state_data)
```

#### 2.2.3.2 National Dataframes

@@

```{r}
full_ntl_data <- income_distribution |> 
  left_join(union_wages, by = "year") |>
  left_join(state_to_ntl_adjusted, by = "year") |>
  left_join(state_to_ntl_employment, by = "year") |>
  mutate(decade = paste0((year - year %% 10), "s"), .after = year) |>
  mutate(mean_median_income_difference = mean_income - median_income, .after = mean_income) |>
   mutate(median_income_2020_adj = median_income * prop_2020_adj,
         mean_income_2020_adj = mean_income * prop_2020_adj,
         mean_median_income_difference_2020_adj = mean_median_income_difference * prop_2020_adj,
         wage_2020_adj = wage * prop_2020_adj,
         union_wage_2020_adj = union_wage * prop_2020_adj,
         nonunion_wage_2020_adj = nonunion_wage * prop_2020_adj) |> 
  select(-prop_2020_adj)
```

```{r}
ntl_column_list <- c("year", "decade", "mean_income",
                     "mean_income_2020_adj", "median_income",
                     "median_income_2020_adj", "mean_median_income_difference",
                     "mean_median_income_difference_2020_adj", "wage",
                     "wage_2020_adj", "union_wage", "union_wage_2020_adj",
                     "nonunion_wage", "nonunion_wage_2020_adj",
                     "wage_percent_difference", 
                     "wage_percent_difference_adjusted",
                     "employment_count", "union_employment_count", 
                     "collective_bargain_employment_count", "union_type")
full_ntl_data <- full_ntl_data[, ntl_column_list] # reorder columns
```

```{r}
rm(house, income_distribution, minimum_wage, state_to_ntl_adjusted,
   state_to_ntl_employment, union_states, union_wages, state_column_list,
   ntl_column_list)
```

@@

## 2.3 EDA

@@
